Refinement for open automata

### 2021-03-26 2021-Refinement/rapport.tex
Refinement v0c

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{cor}{Corollary}
\newtheorem{prop}{Proposition}

@@

\newcommand\nmm[1]{\(\displaystyle #1\)} % nmm for Nice Math Mode
\newcommand\setR{\mathbb{R}}
\newcommand\setZ{\mathbb{Z}}
\newcommand\setN{\mathbb{N}}
\newcommand\mpar[1]{{\left(#1\right)}}
\newcommand\mbrk[1]{{\left[#1\right]}}
\newcommand\mbrc[1]{{\left\{#1\right\}}}
\newcommand\mdbrk[1]{\left⟦#1\right⟧}
\newcommand\card[1]{{\left|#1\right|}}
\newcommand\psubst[1]{\mbrc{\!\!\mbrc{#1}\!\!}}
\newcommand\subst[2]{\mbrk{#1\middle/#2}}
\newcommand\choice[1]{\left\{\everymath{\displaystyle}%
	\begin{array}{lr}#1\end{array}\right.}
\newcommand\midbar{\,\middle|\,}
\newcommand\mset[2]{\mbrc{#1\midbar #2}}
\newcommand\prel[4]{\mbrk{#2\mathrel{#1}#3\midbar #4}}
\newcommand\subbox[1]{{\makebox[.5\width]{\(\scriptstyle #1\)}}}
\newcommand\bigsymb[2][\Large]{\text{#1\nmm{#2}}}
\newcommand\defnotation{\DOTSB\;{\Colon=}\;}
\newcommand\defobject{\DOTSB\;{\coloneq}\;}
\newcommand\nwedge{\DOTSB\;{\wedge}\;} % when you want to force space
\newcommand\qwedge{\DOTSB\quad{\wedge}\quad}
\newcommand\sat{{\vdash}} % suppress spaces
\newcommand\fvars[1]{{\mathit{vars}\mpar{#1}}}
\newcommand\fterms[1]{{\mathcal{T}}}
\newcommand\fcterms[1]{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}}}
\newcommand\fformulae[1]{{\mathcal{F}}}
\newcommand\fcformulae[1]{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}}}
% \newcommand\fterms[1]{{\mathcal{T}\mpar{#1}}}
% \newcommand\fcterms[1]{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}\mpar{#1}}}
% \newcommand\fformulae[1]{{\mathcal{F}\mpar{#1}}}
% \newcommand\fcformulae[1]{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}\mpar{#1}}}
% \newcommand\fvalues[1]{{\mathcal{P}\mpar{#1}}}
\newcommand\fvalues[1]{{\mathcal{P}}}
\newcommand\OA[6]{\left<#1,#2,#3,#4,#5,#6\right>}
\newcommand\OT[6]{\text{\small\(%
	\setlength\arraycolsep{2pt}\everymath{\displaystyle}%
	\begin{array}{c}%
	#4,#5,#6 \\\hline
	#1\xrightarrow{\raisebox{-1.5pt}[.8\height][0pt]{\makebox[1.4\width]{\(\scriptstyle #3\)}}}#2
	\end{array}\)}}
\newcommand\OTx[4]{\OT{s_{#1}}{s'_{#1 #2}}{\alpha_{#1 #2}}{\beta_{#3 j}^{j \in J'_{#4}}}{g_{#1 #2}}{\psi_{#1 #2}}}
\newcommand\OTg{\OTx{}{}{}{}}

% possible examples for refinement: + <= |, ; <= |

@@

\section{Notations}
Throughout this paper, tuples might be denoted differently depending on what they represent.
Quantification will be used loosely, any previously defined object in a quantification means that the object is constrained to have that previous value, and any undefined object is defined by the quantifier.
For instance \(\forall \mpar{x, y} \in \setZ^2, \exists \mpar{x', y} \in \setR^2, P\) means \(\forall \mpar{x, y}, \exists \mpar{x', y'}, y = y' \wedge P\).

Family of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter will only be used when unambiguous; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\), but \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) must be represented with one of the two first notation.
The disjoint union of two maps \(\varphi \in X^I\) and \(\psi \in Y^J\) with \(I \cap J = \emptyset\) is \(\varphi \uplus \psi \in \mpar{X \cup Y}^{I \uplus J}\).

A function with two parameters \(R \in X \times Y \to Z\) might be noted using the relational notation \(\prel{R}{x}{y}{z}\) when it is relevant for noting \(R\mpar{x, y} = z\) or equivalently \(\mpar{x, y, z} \in R\).

\section{Open Automaton}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra]
An expression algebra \(E\) is a disjoint union \(E \defobject \mathcal{T} \uplus \mathcal{F}\) where \(\mathcal{T}\) is a term algebra and \(\mathcal{F}\) the formulae over \(\mathcal{T}\).
% We also define convenience notations \(\fterms{E} \defnotation \mathcal{T}\) and \(\fformulae{E} \defnotation \mathcal{F}\).
\end{defi}
The terms algebra is arbitrary.
The formulae form at least a propositional logic over terms in \(\mathcal{T}\) and some relations, or equivalently they are at least first order logic formulae without quantifiers but with free variables allowed.
An example of term algebra can be Peano integers (zero, variable and successor function), the formulae associated can use the relations equality, \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) and \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\).
\begin{defi}[Free variables, Closed terms/formulae]
\(\fvars{e}\) is the set of unbound variables in \(e \in E\).
The closed terms are \(\fcterms{E} \defnotation \mset{t \in \fterms{E}}{\fvars{t} = \emptyset}\); \(\fcterms{E} \subset \fterms{E}\).
The closed formulae are \(\fcformulae{E} \defnotation \mset{f \in \fformulae{E}}{\fvars{f} = \emptyset}\); \(\fcformulae{E} \subset \fformulae{E}\).
\end{defi}
More generally a closed expression is an expression without unbound variable.
If the expression algebra has no binding construction then closed terms are terms without variables.
\begin{defi}[Interpretation, Values, Satisfaction, (Parallel) substitution]
We assume that the following functions are given:
\begin{itemize}
\item The interpretations of closed terms, \(\mdbrk{\_} \in \fcterms{E} \to \fvalues{E}\), with \(\fvalues{E}\) a set of interpreted value of terms;
\item The satisfaction of closed formulae, \(\sat \subseteq \fcformulae{E} \to \mbrc{0, 1, ?}\);
\item The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \fterms{E}\), \(e\subst{t}{x}\);
\item The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi \in \fterms{E}^V\), \(e\psubst{\psi}\).
% \item The formula interpretation of any variable assignement \(\psi \in \fterms{E}^V\), \nmm{\mdbrk{\psi} \defnotation \bigwedge_{v \in V} v = \psi\mpar{v}}; \(\mdbrk{\psi} \in \fformulae{E}\).
\end{itemize}
\end{defi}
The set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case \(U \defobject V \setminus \fvars{e}, U \neq \emptyset\), the variables in \(U\) are not substituted.
The substitutions might give a nonsensical expression; for instance terms can be integers and pairs with addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is not a valid term.
This can be guarded with \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) when there is quantification on \(t\) and \(\psi\); in the other cases it is supposed to give a valid expression.
The interpretation of values is supposed to be decidable.
The satisfaction of formulae is not supposed to be decidable, however it is better if it is ;).
We are not interested in how \(\fvalues{E}\) is defined.
Whether it is predefined or caracterised by \(\fvalues{E} \defnotation \mdbrk{\fcterms{E}}\) is not the point of this definition.
A value \(v\) may be used for keeping a variable state, and then injected in terms for substitution.
While this is correct when \(\fvalues{E} \subseteq \fcterms{E}\), in the other cases this will be used as a shorthand for substitution with any \(t \in \mdbrk{v}^{-1}\).
\begin{defi}
The following notations will be used for convenience:
\begin{itemize}
\item Let \nmm{f \in \fcformulae{E}}, \nmm{\vdash f_0 \defnotation \sat\mpar{f_0} = 1} and \nmm{\nvdash f_0 \defnotation \sat\mpar{f_0} = 0};
\item Let \nmm{f \in \fformulae{E}, \sigma \in \fvalues{E}^V}, \\
	\nmm{\sigma \vdash f \defnotation \exists \varphi \in \fvalues{E}^{\fvars{f} \setminus V}, f\psubst{\sigma \uplus \varphi} \in \fcformulae{E} \nwedge \vdash f\psubst{\sigma \uplus \varphi}}.
\end{itemize}
\end{defi}
The first notation is quite classical when satisfaction is defined as a relation rather than a function.
The second notation is the satisfiability of formulae under a (partial) valuation.
% \begin{defi}[Formulas combination, Expression algebra combination]
% The propositional logic built without existential quantifiers on top of two propositional logics \(\fformulae{E_1}\) and \(\fformulae{E_2}\) is \(\fformulae{E_1} + \fformulae{E_2}\).
% By extension \(E_1 + E_2 \defnotation \mpar{\fformulae{E_1} + \fformulae{E_2}} \uplus \mpar{\fterms{E_1} \cup \fterms{E_2}}\).
% \end{defi}

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
An open automaton is a tuple \(\OA{S}{i}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\varphi \in \fvalues{E}^V\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, \Quentin{only \(J\) is required to be finite.}{Is \(S\) also required to be finite?}
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in A\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in A\) the action labels of the holes, \(v_\alpha, v_{\beta j}^{j \in J'}\) fresh distinct variables, \(g \in \fformulae{E}\) the guard and \(\psi \in \fterms{E}^V\) the new value of the variables.

\Quentin{The set of action labels \(A\) is arbitrary and implicit.}{For the moment}
To simplify formulae, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.
The guard and new values of the variables can use any defined variable: \nmm{\bigcup_{v \in V} \fvars{\psi\mpar{v}} \cup \fvars{g} \subseteq V \cup \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'}}.
\end{defi}
Now we can define some utilitary functions:
\begin{defi}[Guard, Variables, Out-transition, In-transition]
\(\mathrm{OT}\mpar{s}\) are called the out-transitions of \(s\) and \(\mathrm{IT}\mpar{s}\) and its in-transitions.
\begin{align*}
	\mathit{guard}\mpar{\OTg} & \defnotation g & \mathit{}
	\fvars{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} & \defnotation \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'} \\
	\mathrm{OT}\mpar{s} & \defnotation \mbrc{\OTg \in T} &
	\mathrm{IT}\mpar{s'} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of an open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma \in \fvalues{E}^V\), it cannot perform transitions \nmm{t \defobject \OTg \in T} such that \(\nvdash g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma \uplus \varphi}\) for some \(\varphi \in \fvalues{E}^\fvars{t}\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi} % Examples
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called composition.
\begin{defi}[Composition of open automata]
The composition of \(\mathrm{OA}_c \defobject \OA{S_c}{i_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(\mathrm{OA}_p \defobject \OA{S_p}{i_p}{V_p}{\varphi_p}{J_p}{T_p}\) is:
\begin{align*}
	\mathrm{OA}_p\subst{\mathrm{OA}_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(\mathrm{OA}\psubst{\mathrm{OA}_j^{j \in J}}\).
\end{defi}
The actions emitted when \(\mathrm{OA}_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(\mathrm{OA}_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(\mathrm{OA}_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(\mathrm{OA}_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi} % TODO: Examples
\end{exi}

\section{What is a refinement for Open Automata} % TODO
% Re-read what refinement are used for
There are several properties that we may want from a refinement relation.
Depending on these properties, several kind of refinement may be used.
For example if we are interested in being able to produce the same sequence of action we may want to use trace set inclusion as a refinement.
% simulation refinement, hole type refinement, control refinement, (meet semi)lattice refinement (meet=handshake a <= b <=> a = a || b, join=non-det choice), action refinement

>>>>

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage[hmargin=0.5in,marginparwidth=1.5in,includemp]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}
\newtheorem{cor}{Corollary}
\newtheorem{prop}{Proposition}

@@

\newcommand\nmm[1]{\(\displaystyle #1\)} % nmm for Nice Math Mode
\newcommand\setR{\mathbb{R}}
\newcommand\setZ{\mathbb{Z}}
\newcommand\setN{\mathbb{N}}
\newcommand\mpar[1]{{\left(#1\right)}}
\newcommand\mbrk[1]{{\left[#1\right]}}
\newcommand\mbrc[1]{{\left\{#1\right\}}}
\newcommand\mdbrk[1]{\left⟦#1\right⟧}
\newcommand\card[1]{{\left|#1\right|}}
\newcommand\psubst[1]{\mbrc{\!\!\mbrc{#1}\!\!}}
\newcommand\subst[2]{\mbrk{#1\middle/#2}}
\newcommand\choice[1]{\left\{\everymath{\displaystyle}%
	\begin{array}{lr}#1\end{array}\right.}
\newcommand\midbar{\,\middle|\,}
\newcommand\mset[2]{\mbrc{#1\midbar #2}}
\newcommand\subbox[1]{{\makebox[.5\width]{\(\scriptstyle #1\)}}}
\newcommand\bigsymb[2][\Large]{\text{#1\nmm{#2}}} % DeclareMathDelimiter?
\newcommand\defnotation{\DOTSB\;{\Colon=}\;}
\newcommand\defobject{\DOTSB\;{\coloneq}\;}
\newcommand\nwedge{\DOTSB\;{\wedge}\;} % when you want to force space
\newcommand\qwedge{\DOTSB\quad{\wedge}\quad}
\newcommand\fvars[1]{{\mathit{vars}\mpar{#1}}}
\newcommand\terms{{\mathcal{T}}}
\newcommand\cterms{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle \emptyset\)}}}}
\newcommand\formulae{{\mathcal{F}}}
\newcommand\cformulae{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle \emptyset\)}}}}
\newcommand\values{{\mathcal{P}}}
\newcommand\OA[6]{\left<#1,#2,#3,#4,#5,#6\right>}
\newcommand\OT[6]{\text{\small\(%
	\setlength\arraycolsep{2pt}\everymath{\displaystyle}%
	\renewcommand\arraystretch{1}%
	\begin{array}{c}%
	#4,#5,#6 \\\noalign{\color{red}\hrule height 1pt}
	#1\xrightarrow{\raisebox{-1.5pt}[.8\height][0pt]{\makebox[1.4\width]{\(\scriptstyle #3\)}}}#2
	\end{array}\)}}
\newcommand\OTx[4]{\OT{s_{#1}}{s'_{#1 #2}}{\alpha_{#1 #2}}{\beta_{#3 j}^{j \in J'_{#4}}}{g_{#1 #2}}{\psi_{#1 #2}}}
\newcommand\OTg{\OTx{}{}{}{}}

% possible examples for refinement: + <= |, ; <= |
% transitions between sections
% Re-read what refinement are used for

@@

\section{Notations}
Throughout this paper, tuples might be noted differently depending on what they represent.
Quantification will be used loosely, any bound name in a quantification means that the object is constrained to have that previous value, and any new name is bound by the quantifier.
For instance \(\forall \mpar{x, y}, \exists \mpar{x', y}, P\) means \(\forall \mpar{x, y}, \exists \mpar{x', y'}, y = y' \wedge P\).

Family of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter will only be used when unambiguous; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\), but \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) must be represented with one of the two first notation.
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is \(\varphi \uplus \psi: \mpar{I \uplus J} \to \mpar{X \cup Y}\).

\section{Open Automaton}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra, Formulae, Terms]
An expression algebra \(E\) is a disjoint union \(E \defobject \terms \uplus \formulae\) where \(\terms\) is a term algebra and \(\formulae\) the formulae over \(\terms\).
\end{defi}
The terms algebra is arbitrary.
The formulae form a first order logic over \(\terms\) and some relations.
An example of term algebra can be Peano integers (zero, variable and successor function), the formulae associated can use equality relation, \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) relation and \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\) relation.

\begin{defi}[Free variables, Closed terms, Closed formulae]
\(\fvars{e}\) is the set of unbound variables in \(e \in E\).
The terms restricted to variables in \(V\) are \(\terms_V \defnotation \mset{t \in \terms}{\fvars{t} = \emptyset}\); \(\terms_V \subset \terms\).
The formulae restricted to variables in \(V\) are \(\formulae_V \defnotation \mset{f \in \formulae}{\fvars{f} = \emptyset}\); \(\formulae_V \subset \terms\).
The closed terms are terms restricted to variables in \(\emptyset\), \(\cterms\).
The closed formulae are formulae restricted to variables in \(\emptyset\), \(\cformulae\).
\end{defi}
By extension, expressions restricted to variables and closed expressions are also defined.
A closed expression is an expression without unbound variable.
\begin{defi}[Interpretation, Values, Satisfiability, (Parallel) substitution]
We assume that the following functions are given:
\begin{itemize}
\item The interpretations of closed terms, \(\mdbrk{\_}: \cterms \to \values\), with \(\values\) a set of interpreted value of terms;
\item The satisfiability relation on closed formulae, \({\vdash} \subseteq \cformulae\) and unsatisfiability relation \({\nvdash} \subseteq \cformulae\);
\item The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\);
\item The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{itemize}
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case where it is not, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a nonsensical expression; for instance terms can be integers and pairs with addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is not a valid term.
This can be guarded with the atom \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitely be the case for instance when there is quantification on \(t\) and \(\psi\) to simplify notations.

The interpretation of values is supposed to be decidable.
The (un)satisfiability of formulae might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.\footnote{In practise a term algera with only equality is used, and it is has all these properties.}

A value \(v\) may be used for keeping a variable state, and then injected in terms for substitution.
While this is correct when \(\values \subseteq \cterms\), in the other cases this will be used as a shorthand for substitution with any \(t \in \mdbrk{v}^{-1}\).

We are not interested in how \(\values\) is defined.
Whether it is predefined or caracterised by \(\values \defnotation \mdbrk{\cterms}\) is not the point of this definition.

The satisfiability relations will be noted using the classical relational notation \(\vdash f \defnotation f \in {\vdash}\) and \(\nvdash f \defnotation f \in {\nvdash}\).
The usual satisfiability under a partial valuation is noted \nmm{\sigma \vdash f \defnotation \exists \varphi: \mpar{\fvars{f} \setminus V} \to \values, \vdash f\psubst{\sigma \uplus \varphi}} with \(f \in \formulae, \sigma: V \to \values\).

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
An open automaton is a tuple \(\OA{S}{i}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\varphi: V \to \values\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, only \(J\) is required to be finite.
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables in the formulas.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in A\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in A\) the action labels of the holes, \(v_\alpha, v_{\beta j}^{j \in J'}\) fresh distinct variables, \(g \in \formulae\) the guard and \(\psi: V \to \terms\) the new value of the variables.

The set of action labels \(A\) is arbitrary and implicit.
To simplify formulae, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.
The guard and new values of the variables can use any defined variable: \nmm{\bigcup_{v \in V} \fvars{\psi\mpar{v}} \cup \fvars{g} \subseteq V \cup \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'}}.
\end{defi}
Now we can define some utilitary functions:
\begin{defi}[Guard, Variables, Out-transition, In-transition]
\(\mathrm{OT}\mpar{s}\) are called the out-transitions of \(s\) and \(\mathrm{IT}\mpar{s}\) and its in-transitions.
\begin{align*}
	\mathit{guard}\mpar{\OTg} & \defnotation g & \mathit{}
	\fvars{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} & \defnotation \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'} \\
	\mathrm{OT}\mpar{s} & \defnotation \mbrc{\OTg \in T} &
	\mathrm{IT}\mpar{s'} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of an open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma \in V \to \values\), it cannot perform transitions \nmm{t \defobject \OTg \in T} such that \(\nvdash g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma \uplus \varphi}\) for some \(\varphi: \fvars{t} \to \values\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi}
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called composition.
\begin{defi}[Composition of open automata]
The composition of \(\mathrm{OA}_c \defobject \OA{S_c}{i_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(\mathrm{OA}_p \defobject \OA{S_p}{i_p}{V_p}{\varphi_p}{J_p}{T_p}\) is:
\begin{align*}
	\mathrm{OA}_p\subst{\mathrm{OA}_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(\mathrm{OA}\psubst{\mathrm{OA}_j^{j \in J}}\).
\end{defi}
The actions emitted when \(\mathrm{OA}_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(\mathrm{OA}_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(\mathrm{OA}_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(\mathrm{OA}_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi}
\end{exi}

\section{What is a refinement for Open Automata}
There are several properties that we may want from a refinement relation.
Depending on these properties, several kind of refinement may be used.
For example if we are interested in being able to produce the same sequence of action we may want to use trace set inclusion as a refinement.
Here the main properties that we want are related to composition and action refinement.
A suitable kind of refinement relation for this kind of properties is simulation refinement:
A bisimulation has already been proposed for open automata \marginpar{cite an article here} that is compatible with composition.

% THIS WILL BE RE-WRITTEN LATER
Other kind of refinement relation that \Quentin{I may explore}{This is a note for possible path, some may be explored, some may not.} are control refinement/hole refinement\footnote{No good formulation atm, the idea is that \(a \leq b \defnotation sth \wedge J_a \setminus J_b \leq_{ctrl} J_b \setminus J_a\), sth is probably a clause to ensure that they behave the same without holes involved.}, (meet semi)lattice refinement\footnote{meet = handshake, \(a \leq b \defnotation a = a || b\) with sync holes, I think this will be equivalent to hole-identical sim, (join=non-det choice?)}, weak-simulation refinement\footnote{Weak variation for each interesting simulation refinement}, composition-correct refinement\footnote{Basically a stricter variant of simulation-refinement where simulation has to take place the other way for some transitions (no different holes?) in order to be correct wrt composition}.

The important properties we will consider here are:
\begin{description}
\item[Reflexivity:] \(\forall a, a \leq a\);
\item[Transitivity:] \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\item[Preorder:] Reflexivity and transitivity;
\item[Composition completeness:] \(\forall a\, b, a\mbrk{b} \leq a\);
\item[Composition correctness:] \(\forall a\, b, a \leq b \implies \exists c, a =_{FH} b\mbrk{c}\);
\item[Composition context equivalence:] \(\forall a\, b\, c, a \leq b \implies a\mbrk{c} \leq b\mbrk{c}\);
\item[Composition congruence:] \(\forall a\, b\, c, a \leq b \implies c\mbrk{a} \leq c\mbrk{b}\);
\item[Composition compatibility:] \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\mbrk{a} \leq d\mbrk{b}\);
\item[FH-bisimulation compatibility:] \(\forall a\, b\, c, d, a =_{FH} b \wedge c =_{FH} d \wedge a \leq c \implies b \leq d\);
\item[Deadlock reduction:] If \(a \leq b\) then \(a\) does not introduces new deadlocks, noted \(a \sqsubseteq b\) in the following; % TODO: Find a better place to introduce a better definition of deadlock reduction.
\item[Livelock reduction:] If \(a \leq b\) then \(a\) does not introduces new livelocks, noted \(a \preceq b\) in the following; % TODO: Find a better place to introduce a better definition of livelock reduction.
\item[\(\tau\)-stuttering:] .% TODO
\end{description}
``Deadlock reduction" conflicts with ``composition completeness" by disallowing some unwanted cases where an automaton in a hole cannot produce any action that out-transitions expect. % And some other do too
In particular filling a hole with the deadlock automaton (one state without out-transitions) is not considered a refinement for a no deadlock introductive refinement.
``Composition completeness" can be modified into ``Deadlock reduction completeness" to be compatible with ``Deadlock reduction": \(\forall a\, b, a\mbrk{b} \sqsubseteq a \implies a\mbrk{b} \leq a\).
Only the latter will be considered.
Similar issues arise on other properties and can be fixed by changing the prerequisites to insure that the most deadlock-prone composition of automata is composition complete:
\begin{description}
\item[Composition context equivalence modified:] \(\forall a\, b\, c, a \leq b \wedge a\mbrk{c} \leq a \implies a\mbrk{c} \leq b\mbrk{c}\);
\item[Composition congruence modified:] \(\forall a\, b\, c, a \leq b \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq c\mbrk{b}\);
\item[Composition compatibility modified:] \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq d\mbrk{b}\);
\end{description}

On the other hand there is a property that was discussed, that may arise when designing a refinement relation and that is interesting but actually unwanted:
\begin{description}
\item[Daisy equivalence:] \(a\mbrk{D} \leq a \wedge a \leq a\mbrk{D}\), where \(D\) is the single state automaton that can produce any action.
\end{description}
This property state that composing a hole with an automaton which can do anything is not a strict refinement.
While it might seem to be a natural property because this automaton is the closest to the meaning of a hole in term of automaton, there is actually a difference between the two.
Daisy equivalence allows to refine by filling a hole but also by going the other way, that is creating a hole name and putting its hole action to any transition (and possibly keeping any of the unmodified transitions).
The unwanted behaviour arise for instance when two independant holes are filled, then a hole is created and its actions are put where the two other actions were.
This effectively merges independant holes, or equivalently allows to consider that an automaton can be plug simultaneously into several holes which is not part of the composition semantics.
\begin{exi}
\end{exi}
